### 2021-07-18 Manuscript.Rmd
corrected typos, updated supp data

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^,  Stefan Siebert^3^, Mark Howison^4^, Felipe Zapata^5^, Casey W. Dunn^6^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ Department of Molecular and Cellular Biology, University of California, Davis, California, 95616, USA \newline
^4^ Research Improving People's Lives (RIPL), Providence, RI, USA \newline
^5^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^6^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

Siphonophores are complex colonial organisms, consisting of asexually-produced bodies (called zooids) that share a gastrovascular cavity but that are functionally specialized to specific tasks, including feeding, swimming, and reproducing. Though this extreme functional specialization has captivated biologists for generations, the genomic underpinnings of this phenomenon remain poorly understood. We use RNA-seq to investigate gene expression patterns in five zooids, and one specialized tissue (the pneumatophore, a gas filled float), across seven siphonophore species. Analyses of gene expression across genes and species present several challenges, which we address, in particular the identification of comparable homologous genes and appropriate normalization of expression across species.

@@

Within species, we identified hundreds of zooid-specific genes, as well as a number of putative transcription factors showing differential expression in particular zooids and developmental stages. To compare expression across species, we introduce a novel expression metric and new analytical approaches. We found that gene expression patterns tended to be largely consistent in zooids with the same function across species, but also found evidence of large lineage specific shifts in gene expression.

@@

Differential gene expression analyses enable identification of a large number of genes with zooid-specific expression, however comparisons across species are limited in scope and largely qualitative. Classical comparative analyses enable the identification of species/zooid-specific expression patterns within orthologs. Using branch filtering methods, we extended analyses to a larger number of genes, and identified instances of lineage-specific expression, as well as shared expression within zooids and across species.

@@

Third, we use a novel approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. We call this type of classical analysis Speciation Tree Filtering (STF) [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. This limits analyses to the subset of genes that have no evidence of duplication. This is an all or nothing approach, where gene trees or subtrees (clades) are retained if they have one sequence for each species, and discarded if there are multiple sequences in any of the species. Because this is usually such a small fraction of genes, traditional STF analyses discard a large fraction of the data, including many gene families that may be of interest to the investigator.

@@

Using the methods described above, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. Second, we examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Third we compared gene expression across the seven siphonophore species, to identify putative clade/lineage or zooid specific expression patterns.

@@

Finally, in *Agalma elegans* very few significantly differentially expressed genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` were significantly differentially expressed in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` were significantly differentially expressed in gastric palpons (supplementary file 4). All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are significantly differentially expressed in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were differentially expressed in the B palpons. Most of these genes overlapped with those differentially expressed in the gastric palpons relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

@@

**Speciation Tree Filtering**

@@

expression_pull<-function(object){
x<-Fulldataframe %>% dplyr::select(c(sequence_ids, Gasdev, Palmat,Gasmat, Necdev, Pne, Gonmal, Gonfem ))
subset<-Orthologs2[Orthologs2$Species %in% object@species,]
tpm<-object@tpm[rownames(object@tpm) %in% subset$Gene,] %>% as.data.frame()
treatment<-object@treatment %>% stringr::str_replace_all(.," ",".")
species<-object@species %>% stringr::str_replace_all(.," ",".")
col<-paste(treatment,object@individual, sep="_")
col<-paste(rep(species,length(treatment)),col,sep="_")
colnames(tpm)<-col
tpm<- tpm %>% rownames_to_column(., var="Gene")
tpm<-tpm %>%  left_join(Orthologs2)
tpm <- tpm %>% dplyr::select(.,-c(Species,Gene))
return(tpm)
}

@@

@@

(ref:SOFT-plot) Results of Speciation Tree Filtering analyses. Top row: results using TPM10K expression values; Bottom row: results using TPM10K expression ratios. For both analyses of expression variance explained by species and tissue (right-hand column), we identified highly species-variable genes (SVG, red) and highly tissue-variable genes (TVG, blue). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. A. PCA plot of expression values (TPM10K) of `r res %>% nrow()` strict ortholog genes, showing a species dominated clustering pattern. B.  The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using TPM10K expression values. C. PCA plot of expression ratios of `r res2 %>% nrow()` strict ortholog genes, showing a tissue dominated clustering pattern.  D. The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using expression ratios. Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify a set of `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs for four of the best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*), of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following Breschi et al. @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

**Speciation Branch Filtering**

@@

For our Speciation Branch Filtering (SBF) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (see Methods). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication event. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees represent `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values (TPM10K) were mapped to the tips of the gene trees for each zooid/tissue separately, and ancestral values were inferred at internal nodes with maximum likelihood. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches).

As with the STF analyses, we investigated the impact of counting efficiency on our comparative analyses. We used the TPM10K values at the tips and nodes to calculate expression ratios and calculated changes in expression ratios across a branch for pairs of tissues, using the mature gastrozooid values as the denominator. The distribution of expression ratio changes along species-equivalent branches (branches in the gene tree that correspond to equivalent branches in the species tree, and are descended from speciation events) are shown in figure \@ref(fig:changes-branches2). We found that the variance of change across a given species-equivalent branch is considerably higher for raw TPM10K values as compared with ratios of expression (Fig.\@ref(fig:variance-structure), Fig.  \@ref(fig:variance-structure2)). We also identified the number of gene tree branches that had a negative (change <= -2), positive (change >= 2) or neutral (change > -2 and < 2) change (Fig.\@ref(fig:neg-pos)). For each species-equivalent branch, we found more negative and positive branches than neutral branches when we used TPM10K values, while for expression ratios, we found that the majority of branches show no change across the branch. This suggests that as for the STF analyses, counting efficiency has a large impact on expression change and in turn leads to large gene/branch specific signal. For expression ratios, these results indicate that the vast majority of branches show close to 0 (neutral) change across the branch, suggesting that for closely related genes, expression ratios tend not to differ.

@@

grid.arrange(tree_reference, Changes_raw,Changes_ratio,legend4, bottom="Tissue",widths = c(2,2,0.25),layout_matrix = layout4)

@@

@@

With our STF analyses we found that the vast majority of identified orthologs were neither exclusively species-variable or tissue-variable, however we were nevertheless able to identify a subset of either tissue- or species- variable genes. As has been observed for vertebrate organs, we find based on GO terms that identified species-variable genes tend to play a role in basic cellular functions (so-called 'housekeeping genes'), as compared to tissue-variable genes [@@breschi2016gene].

What can we learn from the SBF analyses, using ratios of expression data? It is important to stress that changes in expression ratios cannot tell us about changes in expression *magnitude*. A tissue-variable gene that is either highly or lowly expressed in, for example, nectophores relative to gastrozooids in all species, may show largely neutral changes across all species-equivalent branches. The sign (positive or negative) of change provides an indication of whether expression in the child node is relatively higher in the denominator or the numerator, relative to the parent node. We find overall that most expression ratios show very little change across species-equivalent gene tree branches -- suggesting that gene expression patterns tend to be largely consistent among species. Positive or negative changes across branches, especially very large changes, across branches, represent putative linage specific shifts in expression. Whether these reflect distinct morphological or functional changes in the tissues of a particular species or clade requires further validation -- a difficulty in this system, where the species are difficult to collect and maintain in the lab.

@@

In this study we used phylogenetic comparative methods to investigate gene expression patterns within homologous zooids across species. To do this, we implemented a new approach called Speciation Branch Filtering. Using gene trees, we considered expression evolution within the context of the gene phylogeny, which may differ significantly from the species phylogeny. Where traditional approaches identify strict ortholog genes *before* conducting analyses, we use information from all gene copies within a gene tree regardless of their evolutionary history of duplication or speciation. By filtering our data by branches, we consider more genes than we would be able to with a strict ortholog approach, and additionally, we investigate expression patterns at both the tips and internal nodes of gene phylogenies. Although we focus here on expression following speciation events, it is also possible to compare expression following duplication and speciation events.

@@

With the expansion of functional genomic tools, including RNA-seq and single cell sequencing methods, there is considerable interest in looking not only at how genomic variation gives rise to phenotypic diversity in a single species or organism, but also at how functional genomic variation shapes phenotypic diversity across multiple closely and distantly related species to understand broader evolutionary patterns and processes [@Brawand:2011du; @barbosa2012evolutionary; @breschi2016gene; @cardoso2019gene; @clarke2017evolutionary; @doi:10.1093/gbe/evw155; @merkin2012evolutionary; @necsulea2014evolution; @perry2012comparative;  @Levin:2016bp; @sudmant2015meta; @ma2018comparative; @yang2013organ; @zhang2014comparative; @10.1093/molbev/msz212; @fukushima2020amalgamated]. Many of these analyses have the goal of identifying shared expression patterns among modular biological units (cell type, tissue, organ, zooid) across species, in order to identify commonalities in expression patterns among species. This is of particular interest for medically orientated fields interested in understanding the extent to which expression results can be extrapolated from one model organism to another. Another goal is to identify expression patterns in a particular biological unit that are unique to a particular species or even clade. For these questions, within-species DGE analyses provide the greatest depth, in terms of number of genes investigated, however comparisons between species on the basis of within-species DGE are limited and largely qualitative. Here we use two approaches to investigate expression patterns in a quantitative manner across homologous tissues, STF and SBF. Classic STF analyses enabled the investigation of a smaller number of strict orthologs that vary in a tissue or species specific manner, however our ability to investigate lineage or especially clade specific patterns of expression in a given zooid are more limited. Meanwhile SBF also focuses on a subset of genes, also with specific evolutionary histories (descended from speciation events), but it enables the identification of expression patterns that vary little among genes/species, as well as expression patterns that show strong lineage or clade specific patterns for a given zooid.

@@

**Collecting** \newline
Specimens were collected in the north-eastern Pacific Ocean in Monterey Bay and, in the case of _Physalia physalis_ the Gulf of Mexico. Specimens were collected by remotely operated vehicle (ROV) or during blue-water SCUBA dives. *Physalia* specimens were collected by hand from the beach after being freshly washed on-shore by prevailing winds. Available physical vouchers have been deposited at the Peabody Museum of Natural History (Yale University), New Haven, CT. Specimens were relaxed using 7.5% MgCl~2~ hexahydrate in Milli-Q water at a ratio of 1/3 MgCl~2~ and 2/3 seawater. Zooids were subsequently dissected from the colony and flash frozen in liquid nitrogen. Colonies were cooled to collection temperatures (e.g 4 degrees C for deep sea species) while the dissections took place. Dissections took no longer than 15-20 minutes. In the case of large colonies, the stem was cut and only partial sections of the colony were placed under the microscope at a given time. Each replicate individual represents a genetically distinct colony from the same species. Replicate specimens were of an equivalent colony size, and zooid replicates were also equivalent sizes. Larger zooid types, such as gastrozooids, were sampled as a single zooid, but smaller zooids were pooled. Pooled zooids were of a comparable maturity and sampled from the same location in a single colony.

@@

This work was supported by the National Science Foundation (DEB-1256695, the Waterman Award, and OCE-1829835). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted at the Yale Center for Research Computing (YCRC). We would like to thank Steve Haddock, Lynne Christianson, and the whole crew of the Western Flyer for their help and expertise. Additional thanks are due to Samuel Church for his assistance during a research cruise, and for productive conversations about comparative gene expression methods.

>>>>

# Evolution of gene expression across species and specialized zooids in Siphonophora {-}
Catriona Munro^1,2^, Felipe Zapata^3^, Mark Howison^4^, Stefan Siebert^5^, Casey W. Dunn^6^
\newline
\newline
^1^ Department of Ecology and Evolutionary Biology, Brown University, Providence, RI 02912, USA \newline
^2^ Current address: Collège de France, PSL Research University, CNRS, Inserm, Center for Interdisciplinary Research in Biology, 75005 Paris, France \newline
^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA 90095, USA \newline
^4^ Research Improving People's Lives (RIPL), Providence, RI, USA \newline
^5^ Department of Molecular and Cellular Biology, University of California, Davis, California, 95616, USA \newline
^6^ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT 06520, USA

@@

Siphonophores are complex colonial animals, consisting of asexually-produced bodies (called zooids) that are functionally specialized for specific tasks, including feeding, swimming, and sexual reproduction. Though this extreme functional specialization has captivated biologists for generations, its genomic underpinnings remain unknown. We use RNA-seq to investigate gene expression patterns in five zooids and one specialized tissue (pneumatophore) across seven siphonophore species. Analyses of gene expression across species present several challenges, including identification of comparable expression changes on gene trees with complex histories of speciation, duplication, and loss. Here, we conduct three analyses of expression. First, we examine gene expression within species. Then, we conduct classical analyses examining expression patterns across species.  Lastly, we introduce Speciation Branch Filtering, which allows us to examine the evolution of expression in a phylogenetic framework.

@@

Within and across species, we identified hundreds of zooid-specific and species-specific genes, as well as a number of putative transcription factors showing differential expression in particular zooids and developmental stages. We found that gene expression patterns tended to be largely consistent in zooids with the same function across species, but also some large lineage-specific shifts in gene expression.

@@

Within-species analyses identify the greatest number of zooid-specific genes, but comparisons across species are limited. Traditional across species analyses focus on the tips of the phylogeny, and identify large-scale expression patterns that are zooid or species variable. Unlike other methods, Speciation Branch Filtering offers an evolutionary perspective into specific changes in expression within zooids along all focal branches, including internal branches.

@@

Third, we use a novel approach to investigate the evolution of gene expression on gene trees with complex evolutionary histories. Most evolutionary studies of gene expression focus exclusively on comparing expression values of strict orthologs (i.e., gene lineages related only by speciation events) which are shared across all species. Analyses of strict orthologs are limited to a subset of genes that have no evidence of duplication, and we call this approach Speciation Tree Filtering (STF) [e.g @Brawand:2011du; @cardoso2019gene; @Levin:2016bp; @yang2013organ]. This is an all or nothing approach, where gene trees or subtrees (clades) are retained if they have one sequence for each species, and discarded if there are multiple sequences in any of the species. Because this is usually such a small fraction of genes, traditional STF analyses discard a large fraction of the data, including many gene families that may be of interest to the investigator.

@@

Using the methods described above, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Second, we compared gene expression across species, using STF followed by linear models to identify zooid- and species-variable genes. And third, we used phylogenetic methods and SBF to identify putative clade/lineage or zooid specific expression patterns.

@@

Finally, in *Agalma elegans* very few significantly differentially expressed genes were identified between the B palpon and gastric palpons  (`r dge_sig[[1]]$PalBmatPalmat %>% nrow()` was significantly differentially expressed in B palpons and `r dge_sig[[1]]$PalmatPalBmat %>% nrow()` were significantly differentially expressed in gastric palpons (supplementary file 4). All three genes have no significant blast hit and did not map to any gene trees). Genes were identified that are significantly differentially expressed in B palpons relative to all other zooids (gastric palpons were excluded). Of this, `r Agalma_PalponB_unique %>% length()` genes were differentially expressed in the B palpons relative to all other zooids. Most of these genes overlapped with those differentially expressed in the gastric palpons relative to all other zooids (`r intersect(Agalma_PalponB_unique,Agalma_Palpon_unique) %>% length()` genes).

@@

**Classical analysis of gene expression across species**

@@

Orthologs1<-read_tsv("strict_orthologs/Single_Copy_Orthologue_Sequences/Strict_orthologs.fa")
Orthologs1<-Orthologs1 %>% mutate(Gene=as.character(Gene))

@@

#### For all 7 species

expression_pull3<-function(object){
x<-Fulldataframe %>% dplyr::select(c(sequence_ids, Gasdev, Palmat,Gasmat, Necdev, Pne))
subset<-Orthologs1[Orthologs1$Species %in% object@species,]
tpm<-x[x$sequence_ids %in% subset$Gene,]
#remove na columns
tpm<-tpm %>% select_if(~sum(!is.na(.)) > 0)
species<-object@species %>% stringr::str_replace_all(.," ",".")
tpm<- tpm %>% dplyr::rename_at(vars(-sequence_ids), ~ paste(species,., sep="_"))
tpm<- tpm %>% dplyr::rename(Gene=sequence_ids) %>% mutate(Gene = as.character(Gene))
tpm<-tpm %>% left_join(subset)
tpm <- tpm %>% dplyr::select(.,-c(Species,Gene))
return(tpm)
}

test3<-lapply(e, expression_pull3)
test3<-purrr::reduce(test3,left_join)

rownames(test3)<-test3$Orthogroup
test3<- test3 %>% na.omit()

@@

(ref:SOFT-plot) Results of classical gene expression analyses across species, using Speciation Tree Filtering followed by linear models. Top row (A & B): results using TPM10K expression values; Bottom row (C & D): results using TPM10K expression ratios. For both analyses of expression variance explained by species and tissue (right-hand column), we identified highly species-variable genes (SVG, red) and highly tissue-variable genes (TVG, blue). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. A. PCA plot of expression values (TPM10K) of `r res %>% nrow()` strict ortholog genes, showing a species dominated clustering pattern. B.  The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using TPM10K expression values. C. PCA plot of expression ratios of `r res2 %>% nrow()` strict ortholog genes, showing a tissue dominated clustering pattern.  D. The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using expression ratios. Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify strict orthologs. For all seven species, we identified `r Orthologs1$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r test3 %>% nrow()` orthologs had expression data across all 7 species. In order to increase the number of genes for analysis, we focused on the four best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*). Using data from 4 taxa, we identified `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore.  Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following Breschi et al. @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

**Phylogenetic analysis of gene expression with Speciation Branch Filtering**

@@

For our Speciation Branch Filtering (SBF) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (branch length threshold <2, less than 0.25 branches/tree with default branch lengths, one or more speciation events present in the tree, fewer than 0.3 nodes that have descendant nodes with the same species node ID, that we able to be time calibrated and finally that have a root depth <5). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication event. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees represent `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values (TPM10K) were mapped to the tips of the gene trees for each zooid/tissue separately, and ancestral values were inferred at internal nodes with maximum likelihood. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches).

As with the STF and linear model analyses, we investigated the impact of counting efficiency on our comparative analyses. We used the TPM10K values at the tips and nodes to calculate expression ratios and calculated changes in expression ratios across a branch for pairs of tissues, using the mature gastrozooid values as the denominator. The distribution of expression ratio changes along species-equivalent branches (branches in the gene tree that correspond to equivalent branches in the species tree, and are descended from speciation events) are shown in figure \@ref(fig:changes-branches2). We found that the variance of change across a given species-equivalent branch is considerably higher for raw TPM10K values as compared with ratios of expression (Figs. \@ref(fig:variance-structure) & \@ref(fig:variance-structure2)). We also identified the number of gene tree branches that had a negative (change <= -2), positive (change >= 2) or neutral (change > -2 and < 2) change (Fig.\@ref(fig:neg-pos)). The mean number of species-equivalent branches considered in these analyses are shown in Table \@ref(tab:summarize-branches) -- this value reflects the mean number of branches across each zooid or zooid ratio (due to incomplete sampling, some zooids may have fewer or more branches; as seen in Fig.\@ref(fig:neg-pos)). For each species-equivalent branch, we found more negative and positive branches than neutral branches when we used TPM10K values, while for expression ratios, we found that the majority of branches show no change across the branch. This suggests that as for the STF analyses, counting efficiency has a large impact on expression change and in turn leads to large gene/branch specific signal. For expression ratios, these results indicate that the vast majority of branches show close to 0 (neutral) change across the branch, suggesting that for closely related genes, expression ratios tend not to differ.

@@

grid.arrange(tree_reference,Changes_raw,Changes_ratio,legend4, bottom="Tissue",widths = c(2,2,0.25),layout_matrix = layout4)

@@

```{r summarize-branches, verbose= FALSE, message = FALSE, echo=FALSE, comment=NA, warning=FALSE}

###table

tpm10kbranches<- edges_tidy %>% filter(treatment%in%c("Gasdev", "Gasmat", "Pne", "Necdev", "Palmat")) %>% group_by(S_new,treatment) %>% dplyr::summarise(nbranch=n())  %>% group_by(S_new) %>% summarize("Mean Number of Branches TPM10K" =mean(nbranch)) %>% rename("Species-equivalent branch"=S_new)

ratio_branches <- edges_tidy %>% filter(treatment%in%c("GasdevGasmat", "PneGasmat", "NecdevGasmat", "PalmatGasmat")) %>% group_by(S_new,treatment) %>% dplyr::summarise(nbranch=n())  %>% group_by(S_new) %>% summarize("Mean Number of Branches ratio" =mean(nbranch)) %>% rename("Species-equivalent branch"=S_new)

branches_table <-tpm10kbranches %>% left_join(ratio_branches) %>% kable(caption="Mean number of species-equivalent branches considered in raw TPM10K or expression ratio analyses.",booktabs=TRUE, align = c("c","c","c"), escape = FALSE) %>% kable_styling(font_size = 11) %>%
  kableExtra::kable_styling(latex_options = "hold_position")

branches_table

@@

With our STF and linear model analyses we found that the vast majority of identified orthologs were neither exclusively species-variable or tissue-variable, however we were nevertheless able to identify a subset of either tissue- or species- variable genes. As has been observed for vertebrate organs, we find based on GO terms that identified species-variable genes tend to play a role in basic cellular functions (so-called 'housekeeping genes'), as compared to tissue-variable genes [@@breschi2016gene].

What can we learn from the SBF analyses, using ratios of expression data? It is important to stress that changes in expression ratios cannot tell us about changes in expression *magnitude*. A tissue-variable gene (identified via linear models in classical analyses) that is either highly or lowly expressed in, for example, nectophores relative to gastrozooids in all species, may show largely neutral changes across all species-equivalent branches. The sign (positive or negative) of change provides an indication of whether expression in the child node is relatively higher in the denominator or the numerator, relative to the parent node. We find overall that most expression ratios show very little change across species-equivalent gene tree branches -- suggesting that gene expression patterns tend to be largely consistent among species. Positive or negative changes across branches, especially very large changes, across branches, represent putative linage specific shifts in expression. Whether these reflect distinct morphological or functional changes in the tissues of a particular species or clade requires further validation -- a difficulty in this system, where the species are difficult to collect and maintain in the lab.

@@

In this study we used three different analyses to investigate gene expression patterns within homologous zooids across species, each are complementary and enable different possibilities for biological discovery. Within-species analyses focus on expression among zooids and tissues within a single species -- a very large number of genes identified in the transcriptome or genome can be considered, and the broader across species gene-tree need not necessarily be considered. This approach is especially useful for investigating expression patterns within novel zooids that do not have a clear homolog in other species. The classical across-species analysis as implemented here, using STF followed by linear models, relies on phylogenetic assumptions yet is a non-phylogenetic tip-focused approach where a single gene is identified per species, and where the identified differences in values across tips are due to changes along all branches in the tree. With linear models, we can identify genes with expression values that are common to a particular zooid across all species, or with expression that is specific to a particular species. However the vast majority of genes are neither zooid or species variable, and show more complex patterns of expression. We proposed a third approach, called Speciation Branch Filtering, that gives access to changes in expression across gene phylogenies, which may differ significantly from the species phylogeny.

Where traditional approaches identify strict ortholog genes *before* conducting analyses, SBF uses information from all gene copies within a gene tree regardless of their evolutionary history of duplication or speciation. By filtering our data by branches, we consider expression patterns at both the tips and internal nodes of gene phylogenies. We are thus able to identify shifts in expression leading to tips as well as shifts in expression leading to particular clades. Although we focus here on expression following speciation events, it is also possible to compare expression following duplication and speciation events.

@@

With the expansion of functional genomic tools, including RNA-seq and single cell sequencing methods, there is considerable interest in looking not only at how genomic variation gives rise to phenotypic diversity in a single species or organism, but also at how functional genomic variation shapes phenotypic diversity across multiple closely and distantly related species to understand broader evolutionary patterns and processes [@Brawand:2011du; @barbosa2012evolutionary; @breschi2016gene; @cardoso2019gene; @clarke2017evolutionary; @doi:10.1093/gbe/evw155; @merkin2012evolutionary; @necsulea2014evolution; @perry2012comparative;  @Levin:2016bp; @sudmant2015meta; @ma2018comparative; @yang2013organ; @zhang2014comparative; @10.1093/molbev/msz212; @fukushima2020amalgamated]. Many of these analyses have the goal of identifying shared expression patterns among modular biological units (cell type, tissue, organ, zooid) across species, in order to identify commonalities in expression patterns among species. This is of particular interest for medically orientated fields interested in understanding the extent to which expression results can be extrapolated from one model organism to another. Another goal is to identify expression patterns in a particular biological unit that are unique to a particular species or even clade. For these questions, within-species DGE analyses provide the greatest depth, in terms of number of genes investigated, however comparisons between species on the basis of within-species DGE are limited and largely qualitative. Here we use two approaches to investigate expression patterns in a quantitative manner across homologous tissues, STF and SBF. Classic STF analyses in conjunction with linear models enabled the investigation of a smaller number of strict orthologs that vary in a tissue or species specific manner, however our ability to investigate lineage or especially clade specific patterns of expression in a given zooid are more limited. Meanwhile SBF also focuses on a subset of genes, also with specific evolutionary histories (descended from speciation events), but it enables the identification of expression patterns that vary little among genes/species, as well as expression patterns that show strong lineage or clade specific patterns for a given zooid.

@@

**Collecting** \newline
Specimens were collected in the north-eastern Pacific Ocean in Monterey Bay and, in the case of _Physalia physalis_ the Gulf of Mexico. Specimens were collected by remotely operated vehicle (ROV) or during blue-water SCUBA dives. *Physalia* specimens were collected by hand from the beach after being freshly washed on-shore by prevailing winds. Available physical vouchers have been deposited at the Peabody Museum of Natural History (Yale University), New Haven, CT. Specimens were relaxed using 7.5% MgCl~2~ hexahydrate in Milli-Q water at a ratio of 1/3 MgCl~2~ and 2/3 seawater. Zooids were subsequently dissected from the colony and flash frozen in liquid nitrogen. Colonies were cooled to collection temperatures (e.g 4 degrees C for deep sea species) while the dissections took place. Dissections took no longer than 15-20 minutes. In the case of large colonies, the stem was cut and only partial sections of the colony were placed under the microscope at a given time. Each replicate individual represents a genetically distinct colony from the same species. Replicate specimens were of an equivalent colony size, and zooid replicates were also equivalent sizes. Larger zooid types, such as gastrozooids, were sampled as a single zooid, but smaller zooids were pooled. Pooled zooids were of a comparable maturity and sampled from the same location in a single colony. Sampling data, including time, date, depth, and voucher ID, can be found in supplementary file 10.

@@

This work was supported by the National Science Foundation (DEB-1256695, the Waterman Award). CM was also supported in part by a RI-EPSCoR Fellowship, NSF EPS-1004057. Analyses were conducted with computational resources and services at the Center for Computation and Visualization at Brown University, supported in part by the NSF EPSCoR EPS-1004057 and the State of Rhode Island. Analyses were also conducted at the Yale Center for Research Computing (YCRC). We would like to thank Steve Haddock, Lynne Christianson, and the whole crew of the Western Flyer for their help and expertise. Additional thanks are due to Samuel Church for his assistance during a research cruise, and for productive conversations about comparative gene expression methods.
